---
title: "Data Science Final Project"
author: "Gabrielle Young"
date: "PSY494/598 Fall 2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

# Introduction

#### Big Research question 
Does meditation influence an individual’s selective attention?

#### Source
The dataset I will be using corresponds to this journal article:
Heino, M. T. (2022). Cognitive Dynamics of a single subject: 1428 Stroop tests 
and other measures in a mindfulness meditation context over 2.5 years. Journal 
of Open Psychology Data, 10. https://doi.org/10.5334/jopd.51 
I found this dataset on Open Science Framework. 

#### Description
The data was collected independently by a 33-year old male doctoral student. 
Each row represents one day of data collection over the course of 900 days. 

I am particularly interested in the following measured variables:

- Stroop_congruent_ pre_meditation: Stroop test consisted of 5 congruent
(e.g. the word "blue" in blue letters) and 15 incongruent trials. This is the 
average response time (in units of 1/100 seconds) to the congruent trials before
doing the meditation.
- Stroop_incongruent_ pre_meditation: Stroop test consisted of 5 congruent and 
15 incongruent (e.g. the word "yellow" in red letters) trials. This is the 
average response time (in units of 1/100 seconds) to the incongruent trials 
before doing the meditation.
- Stroop_congruent_ post_meditation: Stroop test consisted of 5 congruent (e.g.
the word "blue" in blue letters) and 15 incongruent trials. This is the average 
response time to the congruent trials after doing the meditation.
- Stroop_incongruent_ post_meditation: Stroop test consisted of 5 congruent and 
15 incongruent (e.g. the word "yellow" in red letters) trials. This is the 
average response time to the incongruent trials after doing the meditation.
- Meditation_randomised_walking: Meditation was randomised to be performed as
walking meditation. 1 for yes, 0 for no.
- Meditation_randomised_sitting: Meditation was randomised to be performed as 
sitting meditation. 1 for yes, 0 for no.
- Meditation_randomised_standing: Meditation was randomised to be performed as 
standing meditation. 1 for yes, 0 for no.
- Clarity_pre_meditiation: Answer, on a scale of 1-10 in intervals of 0.25, to
the question "How clear do you feel right now".
- Clarity_post_meditation: Answer, on a scale of 1-10 in intervals of 0.25, to 
the question "How clear do you feel right now".
- Calm_pre_meditation: Answer, on a scale of 1-10 in intervals of 0.25, to the 
question "How calm do you feel right now".
- Calm_post_meditation: Answer, on a scale of 1-10 in intervals of 0.25, to the
question "How calm do you feel right now".

```{r load-packages, include = FALSE}
library(dplyr)
library(readr)
library(tidyverse)
library(tidymodels)
library(lubridate)
library(stringr)
library(ggplot2)
```

```{r data-preview, include = FALSE}
cognitive_dynamics_heino <- read_csv("cognitive_dynamics_heino.csv")
glimpse(cognitive_dynamics_heino)
```

\newpage

# Research Question 1  
## Does meditation moderate an individual’s selective attention over time? 

### How did I come up with this question?
- Previous research has indicated that meditation has the potential to 
improve selective attention
- Attention Control Theory: meditation allows for one to focus on specific 
stimuli while ignoring distractions

### Why should you care?
- Most psychological research is conducted using larger sample sizes
- This question puts this theory to the test using a single individual
- Incorporate meditation as a method to instill focus
- Potential long-term positive consequences

```{r clean-data-line-graph, echo = FALSE}
cognitive_dynamics_heino$date <- str_replace_all(cognitive_dynamics_heino$date, "\\.", "/")

cognitive_dynamics_heino$date <- as.Date(cognitive_dynamics_heino$date, format = "%d/%m/%Y")

cogdynrq1 <- cognitive_dynamics_heino %>%
  select(stroop_congruent_pre_meditation, stroop_congruent_post_meditation, 
         stroop_incongruent_pre_meditation, stroop_incongruent_post_meditation,
         date) %>% 
  drop_na() %>% 
  pivot_longer(cols = c("stroop_congruent_pre_meditation", 
                        "stroop_incongruent_pre_meditation", 
                        "stroop_congruent_post_meditation", 
                        "stroop_incongruent_post_meditation"), 
               names_to = "Stroop_Test",
               values_to = "Time")
#create line graph
stroop_labels <- c(
  "stroop_congruent_pre_meditation" = "Pre-Meditation Congruent",
  "stroop_congruent_post_meditation" = "Post-Meditation Congruent",
  "stroop_incongruent_pre_meditation" = "Pre-Meditation Incongruent",
  "stroop_incongruent_post_meditation" = "Post-Meditation Incongruent"
)

cogdynrq1$Stroop_Test <- factor(cogdynrq1$Stroop_Test, levels = c(
 "stroop_congruent_pre_meditation", 
                        "stroop_congruent_post_meditation", 
                        "stroop_incongruent_pre_meditation", 
                        "stroop_incongruent_post_meditation"
))

cogdynrq1 %>% 
  ggplot(aes(x = date, y = Time, color = Stroop_Test, group = Stroop_Test)) +
  geom_line() +
  scale_x_date(date_labels = "%d %b %Y", date_breaks = "2 months", limits = 
                 c(as.Date("2014-08-01"), max(cogdynrq1$date))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        text = element_text(size = 8), strip.text = element_blank()) +
  facet_wrap(~Stroop_Test) +
  scale_color_manual(values = c("#071e26", "#7BB1FB", "#6D1818", "#FB7B7B"), 
  labels = stroop_labels) +  ylim(c(0,20)) +
  labs(title = "Time to Complete Stroop Test Pre/Post Meditation", x = "Date")
```

### Let's disect this:

- August of 2014-February 2017
- No change in selective attention on the Stroop test over time
- These results are consistent regardless of the type (pre, post, congruent, 
or incongruent). 


```{r average-months, echo = FALSE}
cogdynrq1averages <- cogdynrq1
# Convert the date column to a Date object
cogdynrq1averages$date <- as.Date(cogdynrq1averages$date, format = "%y-%m-%d")

# Extract month and year from the date
cogdynrq1averages <- cogdynrq1averages %>%
  mutate(year_month = format(date, "%Y-%m"))

# Group by year_month and mood and calculate the average
averages <- cogdynrq1averages %>%
  group_by(year_month, Stroop_Test) %>%
  summarise(average_monthly_time = mean(Time)) %>% filter(!year_month == "15-03") 

averages %>% 
  ggplot(aes(x = year_month, y = average_monthly_time, 
             color = Stroop_Test, group = Stroop_Test)) +
  geom_line() + theme_minimal() +
  scale_color_manual(values = c("#071e26", "#7BB1FB", "#6D1818", "#FB7B7B"), 
  labels = stroop_labels) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        text = element_text(size = 8)) + labs(title = "Average Response Time", 
                                         subtitle = "By Year-Month", x = "Date",
                       y = "Average Response Time (in units of 1/100 seconds)")
```


### Let's disect this 

- There is a steady decline in average Stroop test response times
- Trend is present across all Stoop Test types
- We might infer this is due to consistent meditation over the months

```{r build-a-model, echo = FALSE}
cogdyn2 <-cognitive_dynamics_heino

model3_congruent_post <- lm(stroop_congruent_post_meditation ~ date, data = cogdyn2)
model4_incongruent_post <- lm(stroop_incongruent_post_meditation ~ date, data = cogdyn2)
model1_congruent_pre <- lm(stroop_congruent_pre_meditation ~ date, data = cogdyn2)
model2_incongruent_pre <- lm(stroop_incongruent_pre_meditation ~ date, data = cogdyn2)
```

\newpage
# Model 1 - Stroop Test Congruent Pre-meditation
```{r model1, echo = FALSE}
summary(model1_congruent_pre)$coefficients
summary(model1_congruent_pre)$r.squared
```
- weak, negative correlation
- small r-squared value

# Model 2 - Stroop Test Incongruent Pre-meditation
```{r model2, echo = FALSE}
summary(model2_incongruent_pre)$coefficients
summary(model2_incongruent_pre)$r.squared
```
- weak, negative correlation
- small r-squared value

# Model 3 - Stroop Test Congruent Post-meditation
```{r model3, echo = FALSE}
summary(model3_congruent_post)$coefficients
summary(model3_congruent_post)$r.squared
```
- weak, negative correlation
- small r-squared value

# Model 4 - Stroop Test Incongruent Post-meditation
```{r model4, echo = FALSE}
summary(model4_incongruent_post)$coefficients
summary(model4_incongruent_post)$r.squared
```
- weak, negative correlation
- small r-squared value


### Let's disect this
- linear model to show the relationship between time taken on a test based on the date
- decrease in time taken to complete stroop test over time
- relationship is prevalent regardless of test type and time of meditation
- all of the relationships between the type of stroop test and point in time
(date) are weak and negatively correlated

\newpage

# Research Question 2
## How do different types of meditation moderate an individual’s selective attention over time?

### How did I come up with this question?
- Attention regulation theory: discusses that meditation may influence selective
attention
- Sitting meditation may contribute to improvements on selective attention
- little research indicating if walking or standing meditations may affect
selective attention

### Why should you care?
- Compare results across the three (sitting, standing, and walking)
- may be useful in selecting the appropriate type of meditation to directly
promote selective attention 
```{r clean-data-rq2, echo = FALSE}
cogdynrq2 <- cognitive_dynamics_heino %>%
select(stroop_congruent_post_meditation, stroop_incongruent_post_meditation,
       meditation_randomised_walking, meditation_randomised_sitting, 
       meditation_randomised_standing, date) %>% 
  drop_na() %>% 
  pivot_longer(cols = c("meditation_randomised_walking", 
    "meditation_randomised_sitting", "meditation_randomised_standing"), 
               names_to = "Meditation",
               values_to = "Time") %>% filter(!Time == 0) %>% select(!Time)
meditation_labels <- c(
  "meditation_randomised_walking" = "Walking", 
  "meditation_randomised_sitting" = "Sitting", 
  "meditation_randomised_standing" = "Standing"
)
```


```{r congruent-performance, echo = FALSE}
cogdynrq2 %>% 
  ggplot(aes(x = date, y = stroop_congruent_post_meditation, color = Meditation, group = Meditation)) +
  geom_point() + geom_smooth() + 
  scale_x_date(date_labels = "%d %b %Y", date_breaks = "2 months", limits = 
                 c(as.Date("2014-08-01"), max(cogdynrq1$date))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        text = element_text(size = 8), strip.text = element_blank()) +
  facet_wrap(~Meditation) +
  scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107"), 
  labels = meditation_labels) +
  labs(title = "Performance Time on Stroop Congruent Based on Meditation Type", 
       subtitle = "Post-Meditation", 
       color = "Meditation Type", x = "Date", y = "Time (in units of 1/100 seconds)")
```

### Let's disect this
- decrease in stroop congruent test completion times across standing and walking 
meditations
- sitting meditation completion times decrease, then level out over time
- potentially a larger increase in selective attention over time for 
standing and walking meditations compared to sitting meditations 

```{r incongruent-performance, echo = FALSE}
cogdynrq2 %>% 
  ggplot(aes(x = date, y = stroop_incongruent_post_meditation, 
             color = Meditation, group = Meditation)) +
  geom_point() + geom_smooth() + 
  scale_x_date(date_labels = "%d %b %Y", date_breaks = "2 months", limits = 
                 c(as.Date("2014-08-01"), max(cogdynrq1$date))) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        text = element_text(size = 8), strip.text = element_blank()) +
  facet_wrap(~Meditation) +
  scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107"), 
  labels = meditation_labels) + ylim(c(0,20)) +
  labs(title = "Performance Time on Stroop Incongruent Based on Meditation Type", 
       subtitle = "Post-Meditation", 
       color = "Meditation Type", x = "Date", y = "Time (in units of 1/100 seconds)")
```

### Let's disect this
- incongruent plot displays similar trends to the previous plot
- decrease in stroop congruent test completion times across standing and walking
meditations
- sitting meditation completion times decrease, then level out over time
- potentially a larger increase in selective attention over time for 
standing and walking meditations compared to sitting meditations 


\newpage

# Research Question 3
## How does meditation affect average daily mood over time?


### How did I come up with this question?
- Theories behind modulating one’s Default Mode Network (DMN) may result in 
feelings of calmness and clarity
- Modulating one's DMN can be achieved through meditation.

### Why should you care?
- We have all probably heard that meditation is good at clearing one's head
- Is this apparent within an individual’s data?

```{r clean-data-r3, echo = FALSE}
cogdynrq3 <- cognitive_dynamics_heino %>%
select(clarity_pre_meditation, calm_pre_meditation, clarity_post_meditation,
       calm_post_meditation, date) %>% 
  drop_na() %>% 
 mutate(average_clarity = rowMeans(select(., clarity_pre_meditation,
                  clarity_post_meditation), na.rm = TRUE)) %>% 
  mutate(average_calm= rowMeans(select(., calm_pre_meditation, 
                  calm_post_meditation), na.rm = TRUE)) %>% 
  select(!c(clarity_pre_meditation,calm_pre_meditation,
            clarity_post_meditation,calm_post_meditation)) %>%
pivot_longer(cols = c("average_clarity", "average_calm"), 
               names_to = "Mood",
               values_to = "Score")
```

```{r rq3, echo = FALSE}
mood_labels <- c(
  "average_calm" = "Calmness", 
  "average_clarity" = "Clarity"
)
cogdynrq3 %>% 
  ggplot(aes(x = date, y =Score , 
             color = Mood, group = Mood)) +
  geom_point() + geom_smooth() + 
  scale_x_date(date_labels = "%d %b %Y", date_breaks = "2 months", limits = 
                 c(as.Date("2014-07-01"), max(cogdynrq1$date))) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        text = element_text(size = 8), strip.text = element_blank()) + facet_wrap(~Mood) +
  scale_color_manual(values = c("#D81B60", "#1E88E5"), labels = mood_labels) + 
  labs(title = "Average Mood Rating over Time", x = "Date", y = 
         "Average Mood Rating (1-10)", color = "Average Mood")
```


### Let's disect this
- decrease in variability over time between average mood ratings
- may be as a result of consistent meditation
- spread is lessened over time

```{r average-monthly, echo = FALSE}
# Convert the date column to a Date object
cogdynrq3$date <- as.Date(cogdynrq3$date, format = "%y-%m-%d")

# Extract month and year from the date
cogdynrq31 <- cogdynrq3 %>%
  mutate(year_month = format(date, "%Y-%m"))

# Group by year_month and mood and calculate the average
result <- cogdynrq31 %>%
  group_by(year_month, Mood) %>%
  summarise(average_monthly_score = mean(Score)) %>% filter(!year_month == "15-03") 
```

```{r graph-average-monthly, echo = FALSE}
result %>% 
  ggplot(aes(x = year_month, y = average_monthly_score, 
             color = Mood, group = Mood)) +
  geom_point() + geom_smooth() + theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        text = element_text(size = 8)) + scale_color_manual(values = 
  c("#D81B60", "#1E88E5"), labels = mood_labels) + labs(title = "Average Mood over Time", 
                                         subtitle = "By Year-Month", x = "Date",
                                        y = "Average Mood Rating (1-10)")
```

### Let's disect this
- Similar trends as the previous plot
- Little change in moods over time
- This is a single individual's data